Microplatforms: Product platforms as code

### 2018-03-14 paper.tex
Add figures

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\section{\label{sec:introduction}Introduction}

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the platform it runs on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See Fig 1).

Service name, Programming language(s), Programming paradigm(s), Architectural choices, Integration pattern(s), Transport protocols, Authentication, Authorization, Reporting, ETLs, Databases, Caching, Platform libraries, Service dependencies, CI Pipeline dependencies, 3rd party library dependencies, 3rd party service dependencies, Security threat model, License audit, Compliance audit, Capacity plan, Provisioning plan, Cost reporting plan, Monitoring plan, Maintenance process, Backup and Restore process, Secret management, Secret rotation, On-Call schedule, Configuration management, Workflow management, Alerts, Log aggregation, Unhandled failure aggregation, Operations and incident response runbooks, API documentation, Source Code Repository, Humane Service Registry, Service Discovery Registry, Distributed Tracing Registry, Monitoring Dashboard Registry, Build Artifact Repository, CI pipeline(s): Build, Test, Publish, Integration tests, Contract tests, Canary, Deploy, Post deploy tests.

Fig 1. 48 concerns for a single microservice [Sean Treadway, SoundCloud] [https://youtu.be/NX0sHF8ZZgw?t=17m48s]

@@

Typically, this is supported by a number of different environments [https://www.phparch.com/2009/07/professional-programming-dtap-%E2%80%93-part-1-what-is-dtap/] starting from the lower ones, where external dependencies of services are stubbed [http://xunitpatterns.com/Test%20Stub.html] and moving all the way up to the production environment which serves customers. Traditionally, these environments are named, long-lived, and are often reused for the purposes of various different testing activities (e.g. functional, performance, and penetration testing).

@@

A platform supporting a digital product which is available to customers any time of day, seven days a week also has to have the ability to tolerate failure of infrastructure or individual components without failure of the product itself. This is commonly referred to as high-availability [https://en.wikipedia.org/wiki/High_availability] and is generally achieved by running several instances of the same services spread across geographical locations, so that the application can keep working even when an entire data center is lost. This horizontal replication is also used for scaling the application with customer demand [https://en.wikipedia.org/wiki/Autoscaling].

An important consideration for a product platform is the cost of building it and the ongoing maintenance involved, including infrastructure provisioning, maintenance (security patches) and scaling capacity to support the products that need to run on it. The effort involved should not be under-estimated [http://www.winnou.com/saas.pdf] and is the motivation behind most digital product now running on cloud infrastructure. We will discuss the reasons for this cost reduction in more detail in sections 3 and 4.

@@

In section \ref{sec:benefits} we outline the goals a good product platform should aim to support. In section 3 the current approach of using a shared platform is discussed and an alternative (Microplatforms) is suggested in Sections 6, supported by the concept of managing all aspects of it as source code in Section 5. Finally, we describe a reference architecture of an example Microplatform in Appendix A.

\section{\label{sec:benefits}Benefits provided by a product platform}

@@

\section{\label{sec:cost}The cost of a shared product platform}

@@

In large organisations, the team structure is typically much more complex, forcing the platform team to request help from other teams, like, for example, IT infrastructure, IT security, networking, etc. amplifying the slow-down even further. This results in a situation sketched in Fig. 2, where each product team has several dependencies on solution delivery teams, each of which have multiple different customers and often depend on each other as well.

Fig 2. - Team dependency structure with shared solution teams

@@

\section{\label{sec:automation}The rise of automation}

@@

Fortunately, as software engineers, we know the way out of this problem: encoding the repetitive tasks (even the ones that require a specialist) in automation. Simplifying a little bit, as more and more tasks are automated, fewer and fewer people and less specialisation is required [Fig 3].

@@

[Fig 4 - scaling customer service, a - engineers needed, b - customer value per engineer]

With increasing level of automation, the number of engineers necessary to support a product grows significantly slower with growth in the number of customers using that product, compared to more manual solutions [ref to google SRE]. As a corollary, with automation, the customer value delivered grows faster with the number of engineers (as more engineers add even more automation, boosting productivity even more), while without it, adding more engineers relatively quickly brings diminishing returns [Fig 4a]. Automation enables relatively small teams of engineers to run products relying on a complex infrastructure whilst supporting millions of customers [ref WhatsApp].

@@

\section{\label{sec:ascode}Product platforms as Code}

@@

\section{\label{sec:microplatforms}}Microplatforms}

@@

Consider the diagram from Fig 4 [automation v resource cost] again, this time with infrastructure approaches and technologies superimposed on the spectrum (Fig 5).

Fig. 5 - Automation techniques

With the currently available level of automation, it is now entirely possible for a cross-functional team of ten to operate a fully-featured product platform for their application. This level of vertical integration has numerous benefits.

@@

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{subcaption}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
%\usepackage{hyperref}% add hypertext capabilities
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\section{Introduction}
\label{sec:introduction}

When building and delivering a real-world client-server application or service, it is important to consider all the parts that form the product, and without which the product would not operate. Typically we think of the product as two major components - the application itself and the platform it runs on. However, there are many other tools, processes, activities and concerns involved in delivering a Software as a Service (SaaS) product to its users than is covered by the functional features and the runtime environment alone (See \ref{fig:concerns}).

\begin{figure}[b]
        \setlength{\fboxsep}{1em}
        \fbox{
                \begin{minipage}{0.43\textwidth}\raggedright
                        Service name, Programming language(s), Programming paradigm(s), Architectural choices, Integration pattern(s), Transport protocols, Authentication, Authorization, Reporting, ETLs, Databases, Caching, Platform libraries, Service dependencies, CI Pipeline dependencies, 3rd party library dependencies, 3rd party service dependencies, Security threat model, License audit, Compliance audit, Capacity plan, Provisioning plan, Cost reporting plan, Monitoring plan, Maintenance process, Backup and Restore process, Secret management, Secret rotation, On-Call schedule, Configuration management, Workflow management, Alerts, Log aggregation, Unhandled failure aggregation, Operations and incident response runbooks, API documentation, Source Code Repository, Humane Service Registry, Service Discovery Registry, Distributed Tracing Registry, Monitoring Dashboard Registry, Build Artifact Repository, CI pipeline(s): Build, Test, Publish, Integration tests, Contract tests, Canary, Deploy, Post deploy tests. 
                \end{minipage}
                }
                \caption{48 concerns for a single microservice [Sean Treadway, SoundCloud] [https://youtu.be/NX0sHF8ZZgw?t=17m48s]}
        \label{fig:concerns}
\end{figure}

@@

Typically, this is supported by a number of different environments [https://www.phparch.com/2009/07/professional-programming-dtap-\%E2\%80\%93-part-1-what-is-dtap/] starting from the lower ones, where external dependencies of services are stubbed [http://xunitpatterns.com/Test\%20Stub.html] and moving all the way up to the production environment which serves customers. Traditionally, these environments are named, long-lived, and are often reused for the purposes of various different testing activities (e.g. functional, performance, and penetration testing).

@@

A platform supporting a digital product which is available to customers any time of day, seven days a week also has to have the ability to tolerate failure of infrastructure or individual components without failure of the product itself. This is commonly referred to as high-availability [https://en.wikipedia.org/wiki/High\_availability] and is generally achieved by running several instances of the same services spread across geographical locations, so that the application can keep working even when an entire data center is lost. This horizontal replication is also used for scaling the application with customer demand [https://en.wikipedia.org/wiki/Autoscaling].

An important consideration for a product platform is the cost of building it and the ongoing maintenance involved, including infrastructure provisioning, maintenance (security patches) and scaling capacity to support the products that need to run on it. The effort involved should not be under-estimated [http://www.winnou.com/saas.pdf] and is the motivation behind most digital product now running on cloud infrastructure. We will discuss the reasons for this cost reduction in more detail in sections \ref{sec:cost} and \ref{sec:automation}.

@@

In section \ref{sec:benefits} we outline the goals a good product platform should aim to support. In section \ref{sec:cost} the current approach of using a shared platform is discussed and an alternative (Microplatforms) is suggested in section \ref{sec:microplatforms}, supported by the concept of managing all aspects of it as source code in section \ref{sec:ascode}. Finally, we describe a reference architecture of an example Microplatform in Appendix A.

\section{Benefits provided by a product platform}
\label{sec:benefits}

@@

\section{The cost of a shared product platform}
\label{sec:cost}

@@

In large organisations, the team structure is typically much more complex, forcing the platform team to request help from other teams, like, for example, IT infrastructure, IT security, networking, etc. amplifying the slow-down even further. This results in a situation sketched in figure \ref{fig:dependencies}, where each product team has several dependencies on solution delivery teams, each of which have multiple different customers and often depend on each other as well.

\begin{figure}[h]
        \includegraphics[width=0.45\textwidth]{figs/dependencies}
        \caption{Team dependency structure with shared solution teams}
        \label{fig:dependencies}
\end{figure}

@@

\section{The rise of automation}
\label{sec:automation}

@@

\begin{figure}[h]
        \includegraphics[width=0.45\textwidth]{figs/automation}
        \caption{Effect of automation}
        \label{fig:automationn}
\end{figure}

Fortunately, as software engineers, we know the way out of this problem: encoding the repetitive tasks (even the ones that require a specialist) in automation. Simplifying a little bit, as more and more tasks are automated, fewer and fewer people and less specialisation is required (figure \ref{fig:automationn}).

@@

\begin{figure}[h]
        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=1\textwidth]{figs/scaling}
                \caption{engineers per customer}
                \label{fig:engineers-needed}
        \end{subfigure}
        \begin{subfigure}{0.45\textwidth}
                \includegraphics[width=1\textwidth]{figs/customervalue}
                \caption{customer value per engineer}
                \label{fig:value-engineer}
        \end{subfigure}
        \caption{Scaling customer service}
        \label{fig:scaling}
\end{figure}

With increasing level of automation, the number of engineers necessary to support a product grows significantly slower with growth in the number of customers using that product, compared to more manual solutions [ref to google SRE]. As a corollary, with automation, the customer value delivered grows faster with the number of engineers (as more engineers add even more automation, boosting productivity even more), while without it, adding more engineers relatively quickly brings diminishing returns (figure \ref{fig:engineers-needed}). Automation enables relatively small teams of engineers to run products relying on a complex infrastructure whilst supporting millions of customers [ref WhatsApp].

@@

\section{Product platforms as Code}
\label{sec:ascode}

@@

\section{Microplatforms}
\label{sec:microplatforms}

@@

Consider the diagram from figure \ref{fig:automationn} again, this time with infrastructure approaches and technologies superimposed on the spectrum (figure \ref{fig:automation-techniques}). With the currently available level of automation, it is now entirely possible for a cross-functional team of ten to operate a fully-featured product platform for their application. This level of vertical integration has numerous benefits.

@@

\begin{figure}[h]
        \includegraphics[width=0.45\textwidth]{figs/automation-tools}
        \caption{Automation techniques}
        \label{fig:automation-techniques}
\end{figure}
